A Bias-Variance Decomposition of Information Transfer in Multi-Task Learning

### 2020-08-17 arxiv/proof_examples.tex
pf

For the isotropic model in Section \ref{sec_similarity}, we actually have an easier and sharper bound than Theorem \ref{thm_model_shift} as follows.
\begin{lemma}[The isotropic model]\label{prop_model_shift_tight}
		In the setting of Theorem \ref{thm_model_shift}, assume that $\Sigma_1 =\id$,
		$\beta_2$ is a random vector with i.i.d. entries with mean $0$, variance $\kappa^2$ and all moments, and $\beta_1$ is a random vector such that $(\beta_1 - \beta_2)$ is a random vector with i.i.d. entries with mean $0$, variance $d^2$ and all moments. Denote
		$\Delta^\star_{\bias} := \bigbrace{(1 - \hat{v})^2 \kappa^2 + d^2} \bigtr{\Pi(\hat v)}.$
	Then we have with high probability,
		\begin{align*}
			\te(\hat{\beta}_t^{\MTL}) \le \te(\hat{\beta}_t^{\STL})\quad \text{ when: }\ \  & \Delta_{\vari} \ge  (\al_+^2(\rho_1)+\oo(1))\cdot  \Delta^\star_{\bias}, \\
			\te(\hat{\beta}_t^{\MTL}) \ge \te(\hat{\beta}_t^{\STL})\quad \text{ when: }\ \  & \Delta_{\vari} \le (\al_-^2(\rho_1)-\oo(1))\cdot  \Delta^\star_{\bias}.
		\end{align*}
\end{lemma}

\begin{proof}%[Proof of Lemma \ref{prop_model_shift_tight}]
The proof of Lemma \ref{prop_model_shift_tight} is similar to that for Theorem \ref{thm_model_shift}, except that we can replace \eqref{bounddelta-} with a tighter bound. We only describe the main difference.

For any fixed $v\in \R$, $\beta_1 - v\beta_2$ is a random vector with i.i.d. entries with mean $0$ and variance $(1-v)^2\kappa^2 + d^2$. Then using the concentration result, Lemma \ref{largedeviation}, we get that for any constant $\e>0$,
\begin{align}
&\left|\delta_{\bias}(v)- [(1-v)^2\kappa^2 + d^2]\tr (\cal K^\top \cal K)\right| \nonumber\\
&=  \left|(\beta_1 - {v} \beta_2)^\top \cal K^\top \cal K (\beta_1 - {v} \beta_2) - [(1-v)^2\kappa^2 + d^2]\tr (\cal K^\top \cal K)\right| \nonumber\\
&\le p^\e [(1-v)^2\kappa^2 + d^2] \left\{\tr\left[(\cal K^\top \cal K)^2\right]\right\}^{1/2} \lesssim p^{1/2+\e} [(1-v)^2\kappa^2 + d^2],\label{anotherbeta}
\end{align}
where we denoted $\cal K:=v\Sigma_2^{1/2}({v}^2 X_1^{\top}X_1 + X_2^{\top}X_2)^{-1} X_1^{\top}X_1$, and in the last step we used $\|\cal K\|=\OO(1)$ by \eqref{eq_isometric}. Now for $\tr (\cal K^\top \cal K)$, we rewrite it as
\begin{align*}
v^2 \tr \left[ ({v}^2 X_1^{\top}X_1 + X_2^{\top}X_2)^{-1}\Sigma_2({v}^2 X_1^{\top}X_1 + X_2^{\top}X_2)^{-1}(X_1^\top X_1)^2 \right].
\end{align*}
Bounding $(X_1^\top X_1)^2=(Z_1^\top Z_1)^2$ using \eqref{eq_isometric} again, we obtain that
\be\label{anotherwtbeta} \delta^\star_{\bias}(v) \cdot (\al_-^2(\rho_1)-\OO(p^{-1/2+\e})) \le  [(1-v)^2\kappa^2 + d^2]\tr (\cal K^\top \cal K) \le \delta^\star_{\bias}(v) \cdot  (\al_+^2(\rho_1)+\OO(p^{-1/2+\e})),\ee
where
$$\delta^\star_{\bias}(v):=n_1^2 v^2 [(1-v)^2\kappa^2 + d^2]\tr \left[ ({v}^2 X_1^{\top}X_1 + X_2^{\top}X_2)^{-1}\Sigma_2({v}^2 X_1^{\top}X_1 + X_2^{\top}X_2)^{-1}\right].$$
Note that $\delta^\star_{\bias}(v) \sim p$, hence combining \eqref{anotherbeta} and \eqref{anotherwtbeta} we get
\be\label{replaceest}
  \delta^\star_{\bias}(v) \cdot (\al_-^2(\rho_1)-\OO(p^{-1/2+\e})) \le  \delta_{\bias}(v)\le \delta^\star_{\bias}(v) \cdot  (\al_+^2(\rho_1)+\OO(p^{-1/2+\e})).
\ee
Now we can replace the estimate \eqref{bounddelta-} with this stronger estimate, and repeat all the other parts of the proof of Theorem \ref{thm_model_shift} to conclude Lemma \ref{prop_model_shift_tight}. In particular, one can calculate $  \delta^\star_{\bias}(v)$ using Lemma \ref{lem_cov_derivative} and get the $  \Delta^\star_{\bias}(v)$ term, We omit the details.
\end{proof}

>>>>

For the isotropic model in Section \ref{sec_similarity}, we actually have an easier and sharper bound than Lemma \ref{thm_model_shift}. %as follows.
\begin{lemma}[The isotropic model]\label{prop_model_shift_tight}
		In the setting of Lemma \ref{thm_model_shift}, assume that $\Sigma_1 =\id$,
		$\beta_2$ is a random vector that has i.i.d. entries with mean $0$, variance $\kappa^2$ and finite moments up to any order, and $(\beta_1-\beta_2)$ is a random vector that is independent of $\beta_2$ and has i.i.d. entries with mean $0$, variance $d^2$ and finite moments up to any order. Denote
		$\Delta^\star_{\bias} := \bigbrace{(1 - \hat{v})^2 \kappa^2 + d^2} \bigtr{\Pi(\hat v)}$, where $\Pi$ was defined in \eqref{defnpihat}.	Then we have with high probability,
		\begin{align*}
			\te(\hat{\beta}_t^{\MTL}) \le \te(\hat{\beta}_t^{\STL})\quad \text{ when: }\ \  & \Delta_{\vari} \ge  (\al_+^2(\rho_1)+\oo(1))\cdot  \Delta^\star_{\bias} \ , \\
			\te(\hat{\beta}_t^{\MTL}) \ge \te(\hat{\beta}_t^{\STL})\quad \text{ when: }\ \  & \Delta_{\vari} \le (\al_-^2(\rho_1)-\oo(1))\cdot  \Delta^\star_{\bias}\ ,
		\end{align*}
		where $\Delta_{\vari}$ was defined in \eqref{Deltavarv}.
\end{lemma}

\begin{proof}%[Proof of Lemma \ref{prop_model_shift_tight}]
The proof of Lemma \ref{prop_model_shift_tight} is similar to that for Lemma \ref{thm_model_shift}, except that we can replace \eqref{bounddelta-} with a tighter bound. We only describe the main difference in the following proof.

For any fixed $v\in \R$, $\beta_1 - v\beta_2$ is a random vector with i.i.d. entries with mean $0$ and variance $(1-v)^2\kappa^2 + d^2$. Then using the concentration result, Lemma \ref{largedeviation}, we get that for any constant $\e>0$, {\cor(start here)}
\begin{align}
&\left|\delta_{\bias}(v)- [(1-v)^2\kappa^2 + d^2]\tr (\cal K^\top \cal K)\right| \nonumber\\
&=  \left|(\beta_1 - {v} \beta_2)^\top \cal K^\top \cal K (\beta_1 - {v} \beta_2) - [(1-v)^2\kappa^2 + d^2]\tr (\cal K^\top \cal K)\right| \nonumber\\
&\le p^\e [(1-v)^2\kappa^2 + d^2] \left\{\tr\left[(\cal K^\top \cal K)^2\right]\right\}^{1/2} \lesssim p^{1/2+\e} [(1-v)^2\kappa^2 + d^2],\label{anotherbeta}
\end{align}
where we denoted $\cal K:=v\Sigma_2^{1/2}({v}^2 X_1^{\top}X_1 + X_2^{\top}X_2)^{-1} X_1^{\top}X_1$, and in the last step we used $\|\cal K\|=\OO(1)$ by \eqref{eq_isometric}. Now for $\tr (\cal K^\top \cal K)$, we rewrite it as
\begin{align*}
v^2 \tr \left[ ({v}^2 X_1^{\top}X_1 + X_2^{\top}X_2)^{-1}\Sigma_2({v}^2 X_1^{\top}X_1 + X_2^{\top}X_2)^{-1}(X_1^\top X_1)^2 \right].
\end{align*}
Bounding $(X_1^\top X_1)^2=(Z_1^\top Z_1)^2$ using \eqref{eq_isometric} again, we obtain that
\be\label{anotherwtbeta} \delta^\star_{\bias}(v) \cdot (\al_-^2(\rho_1)-\OO(p^{-1/2+\e})) \le  [(1-v)^2\kappa^2 + d^2]\tr (\cal K^\top \cal K) \le \delta^\star_{\bias}(v) \cdot  (\al_+^2(\rho_1)+\OO(p^{-1/2+\e})),\ee
where
$$\delta^\star_{\bias}(v):=n_1^2 v^2 [(1-v)^2\kappa^2 + d^2]\tr \left[ ({v}^2 X_1^{\top}X_1 + X_2^{\top}X_2)^{-1}\Sigma_2({v}^2 X_1^{\top}X_1 + X_2^{\top}X_2)^{-1}\right].$$
Note that $\delta^\star_{\bias}(v) \sim p$, hence combining \eqref{anotherbeta} and \eqref{anotherwtbeta} we get
\be\label{replaceest}
  \delta^\star_{\bias}(v) \cdot (\al_-^2(\rho_1)-\OO(p^{-1/2+\e})) \le  \delta_{\bias}(v)\le \delta^\star_{\bias}(v) \cdot  (\al_+^2(\rho_1)+\OO(p^{-1/2+\e})).
\ee
Now we can replace the estimate \eqref{bounddelta-} with this stronger estimate, and repeat all the other parts of the proof of Theorem \ref{thm_model_shift} to conclude Lemma \ref{prop_model_shift_tight}. In particular, one can calculate $  \delta^\star_{\bias}(v)$ using Lemma \ref{lem_cov_derivative} and get the $  \Delta^\star_{\bias}(v)$ term, We omit the details.
\end{proof}

### 2020-08-17 arxiv/transfer.tex
pf

%We now state several helper lemmas to get estimates on $L(\hat{\beta}_t^{\STL})$ and $L(\hat{\beta}_t^{\MTL})$ for $t=2$. 
The next lemma, which is a folklore result in random matrix theory, helps to determine the asymptotic limit of $\te(\hat{\beta}_t^{\STL})=\sigma^2   \bigtr{(X_2^{\top}X_2)^{-1}\Sigma_2}$ as $p\to \infty$. When the entries of $X_2$ are multivariate Gaussian, this lemma recovers the classical result for the mean of inverse Wishart distribution \cite{anderson1958introduction}. For general non-Gaussian random matrices, it can be obtained using Stieltjes transform method; see e.g., Lemma 3.11 of \cite{bai2009spectral}. Here we shall state a result obtained from Theorem 2.4 in \cite{isotropic}, which gives an almost sharp error bound.
\begin{lemma}\label{lem_minv}
Suppose $X$ satisfies assumption \ref{assm_secA1}. Let $A$ be any $p\times p$ matrix that is independent of $X$. We have that for any constant $\e>0$,
	\be\label{XXA}  \bigtr{(X_2^{\top}X_2)^{-1}A} = \frac{1}{\rho_2 - 1} \frac1p\tr(\Sigma^{-1}A) +\bigo{ \|A\|p^{-1/2+\epsilon}} \ee
with high probability.
\end{lemma}
Note that Lemma \ref{lem_cov_shift} extends Lemma \ref{lem_minv} for a single sample covariance matrix to the sum of two independent sample covariance matrices.

@@

Next, we describe our result for more than two tasks with same features, i.e. $X_i = X$ for any $i$.
This setting is prevalent in applications of multi-task learning to image classification, where there are multiple prediction labels/tasks for every image \cite{chexnet17,EA20}.
\begin{theorem}[Many tasks]\label{thm_many_tasks}
%Suppose $X=Z\Sigma^{1/2}\in \R^{n\times p}$ satisfy Assumption \ref{assm_secA1} with $\rho:=n/p>1$ being some fixed constant. Consider data models  $Y_i = X\beta_i + \varepsilon_i$, $i=1,2,\cdots, t$, where $\e_i\in \R^{n}$ are random vectors with i.i.d. entries with mean zero, variance $\sigma^2$ and all moments as in \eqref{assmAhigh}. Moreover, assume that $X$, $\beta_i$ and $\e_i$ are all independent of each other.
	%Let $n = c \cdot p$.
	%Let $X\in\real^{n\times p}$ and $Y_i = X\beta_i + \varepsilon_i$, for $i = 1,\dots,k$.
%	Consider $t$ data models $Y_i = X\beta_i + \varepsilon_i$, $i=1,2,\cdots, t$, where $X$ has covariance matrix $\Sigma$, and the entries of $\e_i$ are i.i.d. with mean zero and variance $\sigma^2$.
	%that satisfy Assumption \ref{assm_secA2} in the appendix.
	For the setting of $t$ tasks where $X_i = X$, for all $1\le i\le t$,
	let $B^\star := [{\beta}_1,{\beta}_2,\dots,{\beta}_{t}]$ and $U_r\in\real^{t\times r}$ denote the linear model parameters.
	Let $U_r U_r^{\top}$ denote the best rank-$r$ subspace approximation of $(B^{\star})^\top\Sigma B^{\star}$.
	Assume that $\lambda_{\min}({B^{\star}}^\top\Sigma B^{\star})\gtrsim \sigma^2$.
	Let $v_i$ denote the $i$-th row vector of $U_r$.
	There exists a value $\delta = \oo \left( \|B^\star\|^2 + \sigma^2\right)$ such that
	\squishlist
		\item  If	$\left(1 - \norm{v_t}^2 \right)\frac{\sigma^2}{\rho - 1} - \norm{\Sigma (B^{\star} U_r v_t - \beta_t)}^2 > \delta$, then w.h.p $\te(\hat{\beta}_t^{\MTL}) < \te(\hat{\beta}_t^{\STL})$.
		\item If $\left(1 - \norm{v_t}^2\right)\frac{\sigma^2}{\rho - 1} - \norm{\Sigma(B^{\star} U_r v_t - \beta_t)}^2 < -\delta$, then w.h.p. $\te(\hat{\beta}_t^{\MTL}) > \te(\hat{\beta}_t^{\STL})$.
	\squishend
\end{theorem}
Theorem \ref{thm_many_tasks} provides a sharp analysis of the bias-variance tradeoff beyond two tasks.
Specifically, $(1 - \norm{v_t}^2)\sigma^2/(\rho - 1)$ shows the amount of reduced variance and $\norm{\Sigma (B^{\star} U_r v_t - \beta_t)}$ shows the bias of the multi-task estimator.
The proof of \ref{thm_many_tasks} can be found in Appendix \ref{app_proof_many_tasks}.

@@

\textit{The isotropic model.}
	Consider two tasks with isotropic covariances $\Sigma_1 = \Sigma_2 = \id$.
	Each task has sample size $n_1 = \rho_1 \cdot p$ and $n_2 = \rho_2 \cdot p$.
	%And $X_1\in\real^{n_1\times p}, X_2\in\real^{n_2\times p}$ denote the covariates of the two tasks, respectively.
	Assume that for  task two, $\beta_2$ has i.i.d. entries with mean zero and variance $\kappa^2$.
	For the source task, $\beta_1 $ equals $\beta_2$ plus i.i.d. entries with mean $0$ and variance $d^2$.
	The labels are $Y_i = X_i\beta_i + \varepsilon_i$, where $\e_i$ consists of i.i.d. entries with mean zero and variance $\sigma^2$.
	For our purpose, it is enough to think of the order of $d$ being $1/\sqrt{p}$ and $pd^2/\sigma^2$ being constant.
%	We assume that all the random variables have subexponential decay, while keeping in mind that our results can be applied under weaker moments assumptions as shown in Appendix \ref{sec_maintools}.

>>>>

%We now state several helper lemmas to get estimates on $L(\hat{\beta}_t^{\STL})$ and $L(\hat{\beta}_t^{\MTL})$ for $t=2$. 
The next lemma, which is a folklore result in random matrix theory, helps to determine the asymptotic limit of $\te(\hat{\beta}_t^{\STL})=\sigma^2   \bigtr{(X_2^{\top}X_2)^{-1}\Sigma_2}$ as $p\to \infty$. When the entries of $X_2$ are multivariate Gaussian, this lemma recovers the classical result for the mean of inverse Wishart distribution \cite{anderson1958introduction}. For general non-Gaussian random matrices, it can be obtained using Stieltjes transform method; see e.g., Lemma 3.11 of \cite{bai2009spectral}. Here we shall state a result obtained from Theorem 2.4 in \cite{isotropic}, which gives an almost sharp error bound.
\begin{lemma}\label{lem_minv}
Suppose $X$ satisfies assumption \ref{assm_secA1}. Let $A$ be any $p\times p$ matrix that is independent of $X$. We have that for any constant $\e>0$,
	\be\label{XXA}  \bigtr{(X_2^{\top}X_2)^{-1}A} = \frac{1}{\rho_2 - 1} \frac1p\tr(\Sigma^{-1}A) +\bigo{ \|A\|p^{-1/2+\epsilon}} \ee
with high probability.
\end{lemma}
Note that Lemma \ref{lem_cov_shift} extends Lemma \ref{lem_minv} for a single sample covariance matrix to the sum of two independent sample covariance matrices. Note that if we set $n_1=0$ and $n_2=n$, then $a_1 = 0$ and $a_2 = (n_2-p) / n_2$ is the solution to \eqref{eq_a12extra}, and one can see that \eqref{lem_cov_shift_eq} reduces to \eqref{XXA}.

@@

Next, we describe our result for more than two tasks with same features, i.e. $X_i = X$ for any $i$.
This setting is prevalent in applications of multi-task learning to image classification, where there are multiple prediction labels/tasks for every image \cite{chexnet17,EA20}.
\begin{theorem}[Many tasks]\label{thm_many_tasks}
%Suppose $X=Z\Sigma^{1/2}\in \R^{n\times p}$ satisfy Assumption \ref{assm_secA1} with $\rho:=n/p>1$ being some fixed constant. Consider data models  $Y_i = X\beta_i + \varepsilon_i$, $i=1,2,\cdots, t$, where $\e_i\in \R^{n}$ are random vectors with i.i.d. entries with mean zero, variance $\sigma^2$ and all moments as in \eqref{assmAhigh}. Moreover, assume that $X$, $\beta_i$ and $\e_i$ are all independent of each other.
	%Let $n = c \cdot p$.
	%Let $X\in\real^{n\times p}$ and $Y_i = X\beta_i + \varepsilon_i$, for $i = 1,\dots,k$.
%	Consider $t$ data models $Y_i = X\beta_i + \varepsilon_i$, $i=1,2,\cdots, t$, where $X$ has covariance matrix $\Sigma$, and the entries of $\e_i$ are i.i.d. with mean zero and variance $\sigma^2$.
	%that satisfy Assumption \ref{assm_secA2} in the appendix.
	For the setting of $t$ tasks where $X_i = X$, for all $1\le i\le t$,
	let $B^\star := [{\beta}_1,{\beta}_2,\dots,{\beta}_{t}]$ and $U_r\in\real^{t\times r}$ denote the linear model parameters.
	Let $U_r U_r^{\top}$ denote the best rank-$r$ subspace approximation of $(B^{\star})^\top\Sigma B^{\star}$.
	Assume that $\lambda_{\min}({B^{\star}}^\top\Sigma B^{\star})\gtrsim \sigma^2$.
	Let $v_i$ denote the $i$-th row vector of $U_r$.
	There exists a value $\delta = \oo \left( \|B^\star\|^2 + \sigma^2\right)$ such that
	\squishlist
		\item  If	$\left(1 - \norm{v_t}^2 \right)\frac{\sigma^2}{\rho - 1} - \norm{\Sigma^{1/2} (B^{\star} U_r v_t - \beta_t)}^2 > \delta$, then w.h.p $\te(\hat{\beta}_t^{\MTL}) < \te(\hat{\beta}_t^{\STL})$.
		\item If $\left(1 - \norm{v_t}^2\right)\frac{\sigma^2}{\rho - 1} - \norm{\Sigma^{1/2}(B^{\star} U_r v_t - \beta_t)}^2 < -\delta$, then w.h.p. $\te(\hat{\beta}_t^{\MTL}) > \te(\hat{\beta}_t^{\STL})$.
	\squishend
\end{theorem}
Theorem \ref{thm_many_tasks} provides a sharp analysis of the bias-variance tradeoff beyond two tasks.
Specifically, $(1 - \norm{v_t}^2)\sigma^2/(\rho - 1)$ shows the amount of reduced variance and $\norm{\Sigma (B^{\star} U_r v_t - \beta_t)}$ shows the bias of the multi-task estimator.
The proof of \ref{thm_many_tasks} can be found in Appendix \ref{app_proof_many_tasks}.

@@

\textit{The isotropic model.}
	Consider two tasks with isotropic covariances $\Sigma_1 = \Sigma_2 = \id$.
	Each task has sample size $n_1 = \rho_1 \cdot p$ and $n_2 = \rho_2 \cdot p$.
	%And $X_1\in\real^{n_1\times p}, X_2\in\real^{n_2\times p}$ denote the covariates of the two tasks, respectively.
	Assume that for  task two, $\beta_2$ has i.i.d. entries with mean zero and variance $\kappa^2$.
	For the source task, $\beta_1 $ equals $\beta_2$ plus i.i.d. entries with mean $0$ and variance $d^2$.
	The labels are $Y_i = X_i\beta_i + \varepsilon_i$, where $\e_i$ consists of i.i.d. entries with mean zero and variance $\sigma^2$.
	For our purpose, it is enough to think of the order of $d$ being $1/\sqrt{p}$ and $pd^2/\sigma^2$ being constant. The more precise conditions on the relations between $d^2$, $\sigma^2$ and $\kappa^2$ are given in \eqref{choiceofpara}.
%	We assume that all the random variables have subexponential decay, while keeping in mind that our results can be applied under weaker moments assumptions as shown in Appendix \ref{sec_maintools}.
